---
title: "Hands On Exercise 8B"
subtitle: "Visualising Geospatial Point Data"
format: html
date: 06/09/2025
date-format: long
date-modified: last-modified
editor: visual
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

[*Note: Please toggle between 'Show' and 'Hide' above based on your preference for viewing the R codes.*]{.underline}

# 8.1 Overview

Proportional symbol maps (also known as graduate symbol maps) are a class of maps that use the visual variable of size to represent differences in the magnitude of a discrete, abruptly changing phenomenon, e.g. counts of people. Like choropleth maps, you can create classed or unclassed versions of these maps. The classed ones are known as range-graded or graduated symbols, and the unclassed are called proportional symbols, where the area of the symbols are proportional to the values of the attribute being mapped. In this hands-on exercise, you will learn how to create a proportional symbol map showing the number of wins by Singapore Pools’ outlets using an R package called **tmap**.

## 8.1.1 Learning Outcome

By the end of this hands-on exercise, you will acquire the following skills by using appropriate R packages:

-   To import an aspatial data file into R.

-   To convert it into simple point feature data frame and at the same time, to assign an appropriate projection reference to the newly create simple point feature data frame.

-   To plot interactive proportional symbol maps.

# 8.2 Getting Started

Before we get started, we need to ensure that **tmap** package of R and other related R packages have been installed and loaded into R.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

# 8.3 Geospatial Data Wrangling

## 8.3.1 Data

The data set use for this hands-on exercise is called *SGPools_svy21*. The data is in csv file format. The XCOORD and YCOORD columns are the x-coordinates and y-coordinates of SingPools outlets and branches. They are in the [Singapore SVY21 Projected Coordinates System](https://www.sla.gov.sg/sirent/CoordinateSystems.aspx).

## 8.3.2 Data Import and Preparation

The code chunk below uses *read_csv()* function of **readr** package to import *SGPools_svy21.csv* into R as a tibble data frame called *sgpools*.

```{r}
sgpools <- read_csv("data/aspatial/SGPools_svy21.csv")
```

After importing the data file into R, it is important for us to examine if the data file has been imported correctly. The code below shows list() is used to do the job.

```{r}
list(sgpools) 
```

Notice that the *sgpools* data in tibble data frame and not the common R data frame.

## 8.3.3 Creating a sf data frame from an aspatial data frame

The code chunk below converts sgpools data frame into a simple feature data frame by using *st_as_sf()* of **sf** packages

```{r}
sgpools_sf <- st_as_sf(sgpools,                         
                       coords = c("XCOORD", "YCOORD"),                        
                       crs= 3414)
```

Things to learn from the arguments above:

-   The *coords* argument requires you to provide the column name of the x-coordinates first then followed by the column name of the y-coordinates.

-   The *crs* argument required you to provide the coordinates system in epsg format. [EPSG: 3414](https://epsg.io/3414) is Singapore SVY21 Projected Coordinate System. You can search for other country’s epsg code by refering to [epsg.io](https://epsg.io/).

You can display the basic information of the newly created sgpools_sf by using the code chunk below.

```{r}
list(sgpools_sf)
```

The output shows that sgppols_sf is in point feature class. It’s epsg ID is 3414. The bbox provides information of the extend of the geospatial data.

# 8.4 Drawing Proportional Symbol Map

To create an interactive proportional symbol map in R, the view mode of tmap will be used.

The code churn below will turn on the interactive mode of tmap.

```{r}
tmap_mode("view")
```

## 8.4.1 It all started with an interactive point symbol map

The code chunks below are used to create an interactive point symbol map.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
           size = 1,
           col = "black",
           lwd = 1)
```

## 8.4.2 Lets make it proportional

To draw a proportional symbol map, we need to assign a numerical variable to the size visual attribute. The code chunks below show that the variable *Gp1Gp2Winnings* is assigned to size visual attribute.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

## 8.4.3 Lets give it a different colour

The proportional symbol map can be further improved by using the colour visual attribute. In the code chunks below, *OUTLET_TYPE* variable is used as the colour attribute variable.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

## 8.4.4 I have a twin brothers :)

An impressive and little-know feature of **tmap**’s view mode is that it also works with faceted plots. The argument *sync* in *tm_facets()* can be used in this case to produce multiple maps with synchronised zoom and pan settings.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) + 
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

Before you end the session, it is wiser to switch **tmap**’s Viewer back to plot mode by using the code chunk below.

```{r}
tmap_mode("plot")
```
