---
title: "Hands On Exercise 2"
subtitle: "Beyond ggplot2 Fundamentals"
date: 04/22/2025
date-format: long
date-modified: last-modified
---

There are several ggplot2 extensions for creating more elegant and effective statistical graphics. Let's explore these features!

# Getting Started

Besides tidyverse, four R packages will be used.

-   [**ggrepel**](https://ggrepel.slowkow.com/): to provide geoms for ggplot2 to repel overlapping text labels

-   [**ggthemes**](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/): to provide extra themes, geoms, and scales for ‘ggplot2’

-   [**hrbrthemes**](https://r-graph-gallery.com/package/hrbrthemes.html): for typography-centric themes and theme components for ggplot2

-   [**patchwork**](https://r-graph-gallery.com/package/patchwork.html): for preparing composite figure created using ggplot2

To check if these packages have been installed and to load them into your working R environment, run the code below.

```{r}
pacman::p_load(ggrepel, patchwork, ggthemes, hrbrthemes, tidyverse)
```

## Importing Data

The code chunk below imports *exam_data.csv* into R environment by using *read_csv()* function of [**readr**](#0) package. 

```{r}
exam_data <- read_csv("Exam_data.csv")
show_col_types = FALSE
knitr::kable(head(exam_data))
```

# What is ggrepel?

One of the challenge in plotting statistical graph is annotation, especially with large number of data points. Take a look below and see it for yourself.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
knitr::opts_chunk$set()
```

ggrepel is an extension of ggplot2 package which provides geoms for ggplot2 to repel overlapping text.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
knitr::opts_chunk$set()
```

# Themes

ggplot2 comes with eight [built-in themes](https://ggplot2.tidyverse.org/reference/ggtheme.html), they are: `theme_gray()`, `theme_bw()`, `theme_classic()`, `theme_dark()`, `theme_light()`, `theme_linedraw()`, `theme_minimal()`, and `theme_void().`

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_dark() +
  ggtitle("Distribution of Maths Scores") 
```

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
knitr::opts_chunk$set()
```

# Combining Graphs

Let's create three graphs named p1, p2 and p3.

```{r}
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")
knitr::opts_chunk$set()

```

```{r}
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")
knitr::opts_chunk$set()
```

```{r}
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
knitr::opts_chunk$set()
```

***Patchwork*** is a ggplot2 extension designed to combine separate ggplot2 graphs into a single figure. Here's the syntax.

-   “/” operator to stack two ggplot2 graphs

-   “\|” operator to place the plots beside each other

-   “()” operator the define the sequence of the plotting

```{r}
p1 + p2
```

```{r}
(p1 / p2) | p3
```

```{r}
patchwork <- (p1 / p2) | p3
patchwork
```
